For Induction-Induction, Induction is Enough

### 2019-11-12 paper/paper.tex
more small fixes, now looking at related work
finished related work

The elimination principle for the above IIT has two motives and one method
for each constructor. It is an example of a \emph{recursive-recursive}
function (following the nomenclature of \cite{forsberg-phd}) which
means that there are two mutually defined functions, where the type of
the second function depends on the first function. The proof assistant
Agda \cite{norell07thesis} allows the definition of such functions
(even from non-IITs) and it is also currently the only proof assistant
supporting IITs. The elimination principle for $\Con$--$\Ty$ says that
given the following components,
\begin{alignat*}{5}
  & Con^M && : \Con\ra\Set \\
  & Ty^M && : Con^M\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & empty^M && : Con^M\,\mathsf{empty} \\
  & ext^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,A\ra Con^M\,(\mathsf{ext}\,\Gamma\,A) \\
  & U^M && : (\Gamma^M:Con^M\,\Gamma)\ra Ty^M\,\Gamma^M\,(\mathsf{U}\,\Gamma) \\
  & El^M && : (\Gamma^M:Con^M\,\Gamma)\ra\,Ty^M\,(ext^M\,\Gamma^M\,(U^M\,\Gamma^M))\,(\mathsf{El}\,\Gamma)
%  & Pi^M && : (\Gamma^M:Con^M\,\Gamma)(A^M:Ty^M\,\Gamma^M\,A)\ra Ty^M\,(ext^M\,\Gamma^M\,A^M)\ra\,Ty^M\,\Gamma^M\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*}
we obtain two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^M\,\Gamma \\
  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^M\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = empty^M \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = ext^M\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^M\,(\mathsf{elimCon}\,\Gamma) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^M\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}

@@

@@

Streicher \cite{streichersemantics} presents an interpretation of the well-formed
presyntax of a type theory into a categorical model, which is an important
ingredient in constructing an initial model, although he does not present
details on the construction of the term model or its initiality proof.

Voevodsky was interested in constructing initial models of type theories
from presyntaxes. There is a communal project \cite{}

Reduction of classes of inductive types: reduce indexed W-types to
W-types. IR types to indexed types. IITs were introduced by
Forsberg. What he did.  QIITs by Forsberg-lots of people. Our QIIT
paper. Jakob TYPES 2019 abstract.  Jasper Huguin: he didn't do
recursion-recursion. Without UIP.

(Indexed inductive types can be reduced to inductive types using a method
similar to what we use here, preterms and well-formedness predicate.)

Talk about connection to setoidified CwFs (I think these are explained
in some Dybjer paper) and maybe the initiality conjecture.

Kaposi and Kov{\'a}cs defined a type theory of signatures for higher
inductive-inductive types (HIITs)
\cite{kaposi_et_al:LIPIcs:2018:9190}.

Similar technique for other type theories: Streicher
\cite{streichersemantics}. Initiality project, Brunerie \cite{}. These
all use quotients. Closest: Lafont TYPES abstract 2017. Does not use
quotients.

Constructing intrinsic syntax but different techniques: Chapman,
Danielsson, Nuo. These use induction-induction which is our goal.

@@

We work in intensional Martin-Löf type theory, equipped with the following features:
\begin{itemize}
  \item Predicative universes $\Set_i$. We usually omit level indices in this paper.
  \item Dependent functions, notated as $(x : A)\ra B$, or as $A \ra B$ when
    non-dependent. We sometimes omit function arguments, by implicitly
    generalising over variables.
  \item $\Sigma$-types, notated either as $(x : A)\times B$, or as $\sum\limits_{x} B$
        when we want to leave the type of the first projection implicit. We use $A
        \times B$ for non-dependent pairs.
  \item Intensional propositional equality as $t = u$, with $\refl : t = t$. In
    this paper, we informally use extensional notation, and do not write
    explicit transports over equalities. However, we occasionally indicate by
    $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the equalities
    $e_1$,\dots,$e_n$.
  \item Uniqueness of identity proofs (UIP), expressing $(e : t = t)\ra e = \refl$.
  \item Function extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$.

@@

We construct all IITs describable by the theory of IIT signatures through the
term model construction of \cite{Kaposi:2019:CQI:3302515.3290315}. There, it is
shown that from the syntax of the theory of signatures for finitary QIITs, one
can construct any particular finitary QIIT. The idea is that for a signature
$\Gamma$, the initial algebra can be built from sets of terms of the form
$\Tm\,\Gamma\,(\El\,a)$. For example, considering the signature for natural
numbers
\[
  \Gamma := \boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N
\]
we have that $\Tm\,\Gamma\,(\El\,N)$ is the set of natural numbers up to isomorphism,
since the only way to construct these terms is by using $s$ and $z$ from $\Gamma$.

>>>>

The elimination principle for the above IIT has two motives and one method
for each constructor. It is an example of a \emph{recursive-recursive}
function (following the nomenclature of \cite{forsberg-phd}) which
means that there are two mutually defined functions, where the type of
the second function depends on the first function. The proof assistant
Agda \cite{norell07thesis} allows the definition of such functions
(even from non-IITs) and it is also currently the only proof assistant
supporting IITs. The elimination principle for $\Con$--$\Ty$ says that
given the following components,
\begin{alignat*}{5}
  & Con^D && : \Con\ra\Set \\
  & Ty^D && : Con^D\,\Gamma\ra\Ty\,\Gamma\ra\Set \\
  & empty^D && : Con^D\,\mathsf{empty} \\
  & ext^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,A\ra Con^D\,(\mathsf{ext}\,\Gamma\,A) \\
  & U^D && : (\Gamma^D:Con^D\,\Gamma)\ra Ty^D\,\Gamma^D\,(\mathsf{U}\,\Gamma) \\
  & El^D && : (\Gamma^D:Con^D\,\Gamma)\ra\,Ty^D\,(ext^D\,\Gamma^D\,(U^D\,\Gamma^D))\,(\mathsf{El}\,\Gamma)
%  & Pi^D && : (\Gamma^D:Con^D\,\Gamma)(A^D:Ty^D\,\Gamma^D\,A)\ra Ty^D\,(ext^D\,\Gamma^D\,A^D)\ra\,Ty^D\,\Gamma^D\,(\mathsf{Pi}\,\Gamma\,A\,B)
\end{alignat*}
we obtain two functions
\begin{alignat*}{5}
  & \mathsf{elimCon} && : (\Gamma:\Con)\ra Con^D\,\Gamma \\
  & \mathsf{elimTy} && : (A:\Ty\,\Gamma)\ra Ty^D\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}
with the following computation rules.
\begin{alignat*}{5}
  & \mathsf{elimCon}\,\mathsf{empty} && = empty^D \\
  & \mathsf{elimCon}\,(\mathsf{ext}\,\Gamma\,A) && = ext^D\,(\mathsf{elimCon}\,\Gamma)\,(\mathsf{elimTy}\,A) \\
  & \mathsf{elimTy}\,(\mathsf{U}\,\Gamma) && = U^D\,(\mathsf{elimCon}\,\Gamma) \\
  & \mathsf{elimTy}\,(\mathsf{El}\,\Gamma) && = El^D\,(\mathsf{elimCon}\,\Gamma)
\end{alignat*}

@@

The contents of this paper were presented at the TYPES 2019 conference
in Oslo \cite{types}.

@@

Streicher \cite{streichersemantics} presents an interpretation of the well-formed
presyntax of a type theory into a categorical model, which is an important
ingredient in constructing an initial model, although he does not present
details on the construction of the term model or its initiality proof. Our paper an indexed variant of his construction (see Subsection \ref{sec:streicher} for a comparison).

Voevodsky was interested in constructing initial models of type
theories from presyntaxes. Inspired by this, Brunerie \cite{brunerie}
formalised Streicher's proof in Agda for a type theory with $\Pi$,
$\Sigma$, $\mathbb{N}$, identity types and an infinite hierarchy of
universes. He used UIP, function extensionality and quotient types in
his formalisation. Our proof uses a similar approach, but we construct
a much simpler type theory which does not have any computation rules,
hence we can avoid using quotients.

Intrinsic (well-typed) syntax for type theories were constructed
before using IITs \cite{chapman09eatitself}, inductive-recursive types
\cite{nisse,Altenkirch:2014:CO:2631172.2631176} and QIITs
\cite{ttintt}. In this paper we avoid using such general classes of
inductive types, as our goal is to reduce IITs to indexed inductive
types.

Reducing general classes of inductive types to simpler classes has a
long tradition in type theory. Indexed W-types were reduced to W-types
\cite{indexedcont} (using the essentially Streicher's idea of preterms
and a typing predicate), small inductive-recursive types to indexed
W-types \cite{malatasta13smallir}, mutual inductive types to indexed
W-types \cite{mutual}. Quotient inductive-inductive types (and
inductive-inductive types) can be reduced to quotient inductive types
using the reduction of generalised algebraic theories to essentially
algebraic theories \cite{gat}.

Our reduction of IITs to indexed inductive types goes through two
steps: first we construct a concrete QIIT using inductive types, then
we construct all IITs from this particular QIIT. A more direct
approach is proposed by \cite{erasure}: here the initial algebra would
be constructed directly for any IIT signature without going through an
intermediate step.

@@

We work in intensional Martin-Löf type theory, equipped with the following features:
\begin{itemize}
  \item Predicative universes $\Set_i$. We usually omit level indices in this paper.
  \item Dependent functions, notated as $(x : A)\ra B$, or as $A \ra B$ when
    non-dependent. We sometimes omit function arguments, by implicitly
    generalising over variables.
  \item $\Sigma$-types, notated either as $(x : A)\times B$, or as $\sum\limits_{x} B$
        when we want to leave the type of the first projection implicit. We use $A
        \times B$ for non-dependent pairs.
  \item Intensional propositional equality as $t = u$, with $\refl : t = t$. In
    this paper, we informally use extensional notation, and do not write
    explicit transports over equalities. However, we occasionally indicate by
    $e_1,\dots,e_n \#t$ that $t$ is well-typed thanks to the equalities
    $e_1$,\dots,$e_n$.
  \item Uniqueness of identity proofs (UIP), expressing $(e : t = t)\ra e = \refl$.
  \item Function extensionality, expressing $((x : A) \ra f\,x = g\,x) \ra f = g$.
  \item Natural numbers $\mathbb{N}$ with constructors $\mathsf{0}:\mathbb{N}$ and $\mathsf{S}:\mathbb{N}\ra\mathbb{N}$.

@@

We construct all IITs describable by the theory of IIT signatures through the
term model construction of \cite{Kaposi:2019:CQI:3302515.3290315}. There, it is
shown that from the syntax of the theory of signatures for finitary QIITs, one
can construct any particular finitary QIIT. The idea is that for a signature
$\Gamma$, the initial algebra can be built from sets of terms of the form
$\Tm\,\Gamma\,(\El\,a)$. For example, considering the signature for natural
numbers
\[
  \Gamma := \big(\boldsymbol{\cdot}\,\ext\,N:\U\,\ext\,z : \El\,N\,\ext\,s : N \ra \El\,N\big),
\]
we have that $\Tm\,\Gamma\,(\El\,N)$ is the set of natural numbers up to isomorphism,
since the only way to construct these terms is by using $s$ and $z$ from $\Gamma$.
